---
output: github_document
---

<!-- r_cpp.md is generated from r_cpp.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# Notes of C++ and R

The following script aims to review the function of [`Rcpp`](https://mran.microsoft.com/snapshot/2017-04-11/web/packages/Rcpp/vignettes/Rcpp-introduction.pdf). As a sociologist, I always struggle to understand some underlying processes, and I am curious about digging a bit more into this issue!

<div class="alert alert-info">
**WARNING!** All my notes are for macOS users. 
</div>

Pending,

- Deconstruct some `C++` codes for clarity
- Re-write some notes and expand them
- Add some `Rcpp` codes for Social Network Analysis (maybe new script)
- Create a new script for `RcppArmadillo`


## C++ and Rcpp

`C++` is a compiled and statistically typed language. 

According to *Wikipedia*? 

> *C++ is a general-purpose programming language created by Bjarne Stroustrup as an extension of the C programming language, or "C with Classes".*
> `r tufte::quote_footer('[Wikipedia](https://en.wikipedia.org/wiki/C%2B%2B)')`

Why `Rcpp`?

> *Sometimes R code just isn’t fast enough. You’ve used profiling to figure out where your bottlenecks are, and you’ve done everything you can in R, but your code still isn’t fast enough [...] learn how to improve performance by rewriting key functions in C++. This magic comes by way of the [`Rcpp`](http://www.rcpp.org) package (Eddelbuettel and François [2011](https://adv-r.hadley.nz/rcpp.html#ref-Rcpp)) (with key contributions by Doug Bates, John Chambers, and JJ Allaire).*
> `r tufte::quote_footer('[Advanced R](https://adv-r.hadley.nz/rcpp.html)')`


<div class="alert alert-info">
From here I would assume that you have some previous background in the `C++` programming^[If you are a newbie to `C++`, you might have some questions. In [Reddit](https://www.reddit.com/r/learnprogramming/wiki/faq_cpp?utm_source=share&utm_medium=ios_app&utm_name=iossmf) there are some FAQ! If you are curious, you can actually hear Bjarne Stroustrup in [YouTube](https://www.youtube.com/watch?v=JBjjnqG0BP8)].
</div>


## Rcpp and Social Network Analysis

There are very nice packages written in `R` that use `Rcpp` in some part of the codes for social network analysis. Some examples are:

- [lolog](https://github.com/statnet/lolog)
- [multinet](http://multilayer.it.uu.se/jss.pdf)
- [NetSim](https://ethz.ch/content/dam/ethz/special-interest/gess/social-networks-dam/documents/jss_netsim.pdf)
- [backbone](https://github.com/domagal9/backbone)


## Setup

Before using `Rcpp` I would need to install `Xcode` in the app store. For Windows, [`Rtool`](https://cran.r-project.org/bin/windows/Rtools/) have to be installed, and in Linux use a code such as `sudo apt-get install r-base-dev`. 

## Rcpp Introduction

We would now create a `C++` function- However, the most important element in the following code to connect `R` with `C++` is this code `// [[Rcpp::export]]`, which allow exporting the `cumSum` function that is created in the following chunk into `R`. 
```{r engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector cumSum(NumericVector x) {
  int n = x.size();
  NumericVector out(n);
  out[0] = x[0];
  for (int i = 1; i < n; ++i) { 
    
    /*
     ++i: incrementing (adding 1 to a variable)
     --i: decrementing (subtracting 1 from a variable)
     i++: increment after the variable's value is used in the remainder of the expression 
     i--: decrement occurs after the variable's value is used in the remainder of the expression 
    */
    
    out[i] = out[i-1] + x[i];
  }
  return out;
}
```

Same code in `R`:
```{r}
CumSum <- function(x) {
    output <- numeric(length(x))
    output[1] <- x[1]
    for (i in 2:length(x)) {
      output[i] <- output[i-1] + x[i]
    }
    return(output)
  }
```

This is the process that we are calculating with the function in the first iteration $(0-1)+2$, then $(1-1)+3$, and $(3-1)+4....$. Recall that `C++` start from $0$ rather than $1$ (*in C++, vectors indices start at 0*). Now, we have three similar functions. One from `base R`, the other is a `C++` function (called: `cumSum`), and the loop `CumSum` created in `R`. They should give the same results:
```{r, results='hold'}
x <- 1:10
cumsum(x)
cumSum(x)
CumSum(x)
```

Now, we can compare the performance of each function using `microbenchmark`. `Rcpp` should be faster than the `R` functions:
```{r}
library(magrittr)
library(microbenchmark)
x <- 1:1000
microbenchmark(
  native = cumsum(x),
  loop = CumSum(x),
  Rcpp = cumSum(x)
) %>% summary(unit = "ms") %>% knitr::kable(format = "markdown")
```

The speed and ability to cheaply use resources are two of the main features of `C++`. Due that is a compiler and not an interpreter is one of the reasons why is faster than `R`.


Using `Rcpp` we can use `C++` codes directley into `R` (`evalCpp`)
```{r warning=FALSE}
library(Rcpp)

# evaluate simple expression as C++
Rcpp::evalCpp("3 + 2") 

set.seed(42)
Rcpp::evalCpp("Rcpp::rnorm(2)")

# maximum numeric limits of my computer in terms of double
Rcpp::evalCpp("std::numeric_limits<double>::max()")
```

Also, we can add functions from `C++` (`cppFunction`) into the `R` environment. 
```{r}
library(Rcpp)
## Example 1
# create a C++ function to be used in R
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')

# R function with same name as C++ function
add
add(1, 2, 3)

## Example 2
# create a C++ function to be used in R
cppFunction("
            int exampleCpp11(){
            auto x = 10; // guesses type
            return x;
            }", plugins=c("cpp11")) ## C++11 is a version of the standard for the programming language C++

# R function with same name as C++ function
exampleCpp11()

## Other examples
cppFunction("
            int doubleMe(int a){return a+a;}
            ")
doubleMe
a <- 3
doubleMe(a)

cppFunction("
            int f(int a, int b){
            return(a+b);
            }
            ")
f(21, 21)
f(21.0, 21)
f(21.5, 21.5) # !
#f(21.5, "hello, world") # error!

```

There are some issues with `integer` and `double`
Some names that are used in `R`:

- integer numbers: `int`
- Floating point number: `double`
```{r}
## R
# literal numbers are double
x <- 42
storage.mode(x)

# integers needs the L suffix
y <- 42L
storage.mode(y)

z <- as.integer(42)
storage.mode(z)

## C++
library(Rcpp)

# Literal integers are `int`
x <- evalCpp("42")
storage.mode(x)

# Suffix with .0 forces a double
y <- evalCpp("42.0")
storage.mode(y)

# explicit casting with (double)
y <- evalCpp("double(40+2)")
storage.mode(y)

# Beware of the integer division!
# integer division
evalCpp("13/4")

# explicit cast, and hence use of double division
evalCpp("(double)13/4")

```

The last function of the package `Rcpp` is `sourceCpp()` that is the source code in which other functions are built (e.g. `evalCpp` and `cppFunction`). This function builds on and extends `cxxfunction()`. On the features is that can compile a number of functions and create a file from all of them. Also, have some plugins and dependency options (e.g. `RcppArmadillo`,  `RcppEigen`, `RcppSGL`). These plugins can also turn on support for `C++11`, `OpenMp`, and more.

We can also write some `C++` codes directly in RMarkdown adding to the chunk the following line of codes `r engine='Rcpp'`

Now we will create an example, and create a function called `timesTwo` and run the `C++` code as an `R` function from `C++`. The most important element is the code `// [[Rcpp::export]]` which export the following function into `R`. Also, we would ask for `Rcpp` to `source` the function for us. Behind the scenes `Rcpp` creates a wrapper, and then `Rcpp` compiles, links, and loads the wrapper. Notice that `std` is tacking the standard package of `C++` to conduct a second function.
```{r engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

// This is a simple example of exporting a C++ function to R. You can
// source this function into an R session using the Rcpp::sourceCpp
// function (or via the Source button on the editor toolbar). ...

// [[Rcpp::export]]
NumericVector timesTwo(NumericVector x){
  return x * 2;
}

// You can include R code blocks in C++ files processed with source Cpp
// (useful for testing and development). The R code will be automatically
// fun after the compilation.

/*** R
timesTwo(42)
cat("hello!\n")
timesTwo(c(12,48,28))
*/

// [[Rcpp::export]]
double cubed(double x){
  return std::pow(x, 3);
}

// [[Rcpp::export]]
NumericVector cubedVec(NumericVector x){
  return pow(x, 3);
}

```

Actually, if we check once again, we would have available the function into our `R` environment. 
```{r}
timesTwo(42)
cubed(2)
cubedVec(3)
```

Now we would check the speed of a function from `R` and another from `C++`. We would try the Fibonacci sequence
```{r}
# An R function
f <- function(n){
  if(n < 2) return(n)
  return(f(n-1) + f(n-2))
}
sapply(0:10, f)

# A C++ function
Rcpp::cppFunction('int g(int n){
                  if (n < 2) return(n);
                  return(g(n-1) + g(n-2));
                  }
                  ')
sapply(0:10, g)

# Check time! increase exponentially
library(rbenchmark)
benchmark(f(10), f(15), f(20))[,1:4]

# Now we would compare both functions, C++ perform incredible faster!
benchmark(f(20), g(20))[,1:4]
res <- microbenchmark::microbenchmark(f(20), g(20))
res
suppressMessages(microbenchmark:::autoplot.microbenchmark(res))

```


## D. Eddelbuettel tutorial

Some resources:

- [rcpp](http://www.rcpp.org)
- [Rcpp Gallery](https://gallery.rcpp.org)
- [Blog](http://adv-r.had.co.nz/Rcpp.html)
- [LearnCpp](https://www.learncpp.com)
- [cplusplus](http://www.cplusplus.com)
- [Seamless R and C++ Integration with Rcpp](https://www.springer.com/gp/book/9781461468677)
- [D. Eddelbuettel tutorial](https://www.youtube.com/watch?v=57H34Njrns4)
- [RStudio](https://www.mjdenny.com/Rcpp_Intro.html)


In comparison with `R`, `C++` is a compiler and not interpreted. Therefore, we may need to supply 

- *header location* via `-I`
- *library location* via `-L`
- *library* via `--llibraryname`

Then, we would need to provide these items and compile them all together as,

`g++ -I/usr/include -c qnorm_rmath.cpp`
`g++ -o qnorm_rmath qnorm_rmath.o -L/usr/lib -lRmath`

Some notes of the differences between `R` and `C++`:

- `R` is dynamically typed: `x <- 3.14; x <- "foo"` is valid.
- In `C++`, each variable must be declared before first use.
- Common types are `int` and `long` (possible with `unsigned`), `float` and `double`, `bool`, as well as `char`.
- No standard string type, though `std::string` is close.
- All these variables types are scalars which is fundamentally different from `R` where everything is a vector
- `class` and (`struct`) allow the creation of composite types; classes add behaviour to data to form `objects`
- Variables need to be declared, cannot change
- Control structures similar to `R`: `for`, `while`, `if`, `switch`
- Functions are similar too but note the difference in position-only matching, also same function name but different arguments allowed in `C++`
- Pointers and memory management: very different but lots of issues people had with `C` can be avoided via `STL` (which is something `Rcpp` promotes too)
- Sometimes still useful to know that a pointer is...

Comparison between `C++ `: Scalar; and `Rcpp` Vectors

- `int`: IntegerVector
- `double`: NumericVector
- `char[];std::string`: CharacterVector
- `bool`: LogicalVector
- `complex`: ComplexVector

```{r engine='Rcpp'}
#include <Rcpp.h>
// [[Rcpp::export]]
double getMax(Rcpp::NumericVector v){
  int n = v.size(); // vectors are describing
  double m = v[0]; // initialize
  for (int i=0; i<n; i++){
    if (v[i] > m){
      Rcpp::Rcout << "Now" << m << std::endl;
      m = v[i];
    }
  }
  return(m);
}

```

```{r}
getMax(c(20, 201, 18))
```

```{r engine='Rcpp', eval=FALSE}
#include <Rcpp.h>

//[[Rcpp::export]]
Rcpp::NumericVector colSums(Rcpp::NumericMatrix mat){
  size_t cols = mat.cols(); // How big is the matrix?
  Rcpp::NumericVector res(cols); // knowing the size, we could create a vector
  for (size_t i=0; i<cols; i++){ // Then operation from 0 from less of the columns
    res[i] = sum(mat.column(i));
  }
  return(res);
}

```

What we did?

- `NumericMatrix` and `NumericVector` go-to types for matrix and vector operationso n floating point variables
- We prefix with `Rcpp::` to make the namespace explicit
- Accessor functions `.rows()` and `.cols()` for dimensions
- Result vector allocated based on number of columns column
- Function `column(i)` extracts a column, gets a vector, and `sum()` operates on it
- The last `sum()` was internally vectorised, no need to loop over all elements

Another example
```{r engine='Rcpp', eval=FALSE}
#include <Rcpp.h>

//[[Rcpp::export]]
double getMax(Rcpp::NumericVector v){
  int n = v.size(); // vectors are describing
  double m = v[0];
  for (int i=0; i<n; i++){
    if (v[i] > m) {
      Rcpp::Rcout << "Now" << m << std::endl;
      m = v[i];
    }
  }
  return(m);
}

```

```{r}
getMax(c(1,4,3,46))
```

```{r engine='Rcpp', eval=FALSE}
struct Date {
  unsigned int year;
  unsigned int month;
  unsigned int day
};

struct Person {
  char firstname[20];
  char lastname[20];
  struct Date birthday;
  unsigned long id;
};

class Date {
private:
  unsigned int year;
  unsigned int month;
  unsigned int date;
public: 
  void setDate(int y, int m, int d);
  int getDay();
  int getMonth();
  int getYear();
}

```


`C++` has vectors as well: written as `std::vector<T>` where `T` denotes template meaning different *types* can be used to instantiate
```{r}
Rcpp::cppFunction("double getmax2(std::vector<double> v){
  int n = v.size(); // vectors are describing
  double m = v[0]; // initialize
  for(int i=0; i<n; i++){
    if (v[i] > m){
    m = v[i];
    }
  }
}")

getMax(c(4,5,2))

```

- STL vectors are widely used so `Rcpp` supports them
- Very useful to access other `C++` code and libraries
- One caveat: `Rcpp` vectors *reuse* R memory so no copies
- STL vectors have different underpinning so copies
- But not a concern unless you have a) either HUGE data structures, b) or many many calls


```{r}
Rcpp::cppFunction("void setSecond(Rcpp::NumericVector v){
            v[1] = 42; // numeric vector (temporal copy! could not survey)
}")

v <- c(1,2,3); setSecond(v); v # as expected
v <- c(1L, 2L, 3L); setSecond(v); v # different (vector of integers)

```

```{r engine='Rcpp'}
#include <Rcpp.h>

// [[Rcpp::export]]
double getMax(Rcpp::NumericVector v){
  return( max(v));
}

```

```{r}
getMax(c(2,3,4,5))
```

For math operations, it should be used `RcppArmadillo`!

**Packages**

[Check this](https://www.youtube.com/watch?v=SgXVRHqh9l8)

- The standard unit of R code organization.
- Creating packages with Rcpp is easy
  - Create an empty one via `Rcpp.package.skeleton()`
  - Rstudio has the File -> New Project -> Directory Choices -> Package Choices
- The vignette `Rcpp-packages` has fuller details
